---
title: "midterm project"
author: "Ruicong(Timothy) Du"
date: "2025-03-30"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

we first need to load the required libraries and read our dataset
 Load necessary library
```{r}
library(ggplot2)
library(readr)
library(tidymodels)  
library(splines) 
library(mgcv)
library(caret)
library(earth)
library(pdp)
library(gridExtra)
library(plotly)
library(patchwork)
```

Next, we need to transform the R datafile into CSV format and read the datafile
```{r}
load("dat1.RData")
load("dat2.RData")

# Save as CSVs
write.csv(dat1, "dat1.csv", row.names = FALSE)
write.csv(dat2, "dat2.csv", row.names = FALSE)

dat1 = read_csv("dat1.csv",  na = c("NA","","."))
dat2 = read_csv("dat2.csv",na = c("NA","","."))

glimpse(dat1)
glimpse(dat2)
```

Next, we will explore dat1 and perform Exploratory analysis
```{r}
dat1 %>% 
  select(where(is.numeric)) %>% 
  summary()

# Count and proportion tables for categorical variables
dat1 %>% 
  select(gender, race, smoking, diabetes, hypertension) %>% 
  map(~table(.))  

dat1 %>% 
  select(gender, race, smoking, diabetes, hypertension) %>% 
  map(~prop.table(table(.)))  

```

Exploratory Visualizations:

1. Visualize the distribution of log_antibody

```{r}
ggplot(dat1, aes(x = log_antibody)) +
  geom_histogram(bins = 30, fill = "skyblue", color = "white") +
  labs(title = "Distribution of Antibody Levels", x = "log_antibody", y = "Count") +
  theme_minimal()
```

* log_antibody is approximately normally distributed and symmetric.
* No major skewness or extreme outliers. 
* Mean centered around 10.


2. Scatterplot: log_antibody vs. time: See how antibody levels change over time post-vaccination.

```{r}
plot_ly(
  data = dat1,
  x = ~time,
  y = ~log_antibody,
  type = "scatter",
  mode = "markers", marker = list(opacity = 0.3)
) %>%
  add_lines(
    x = ~time,
    y = ~fitted(loess(log_antibody ~ time, data = dat1)),
    line = list(color = "red"),
    name = "LOESS Smooth"
  ) %>%
  layout(
    title = "Antibody Levels Over Time",
    xaxis = list(title = "Time Since Vaccination (days)"),
    yaxis = list(title = "log_antibody")
  )
```

* Antibody levels increase initially, peak around day 75, then gradually decline.

* Clear nonlinear relationship between log_antibody and time

3. Boxplots by categorical variables: To compare antibody levels between different subgroups

Gender
```{r}
p_gender = ggplot(dat1, aes(x = factor(gender), y = log_antibody)) +
  geom_boxplot(fill = "lightblue") +
  labs(title = "Antibody Levels by Gender", x = "Gender (0 = Female, 1 = Male)", y = "log_antibody") +
  theme_minimal()

p_gender
```

* Females (0) tend to have slightly higher median antibody levels than males (1).
* Distributions overlap but differ in central tendency.

Smoking status
```{r}
p_smoking = ggplot(dat1, aes(x = factor(smoking), y = log_antibody)) +
  geom_boxplot(fill = "plum") +
  labs(title = "Antibody Levels by Smoking Status", x = "Smoking (0=Never, 1=Former, 2=Current)", y = "log_antibody") +
  theme_minimal()

p_smoking
```
* Never and former smokers have similar distributions, with slightly higher medians.
* Current smokers tend to have slightly lower antibody levels.
* Smoking status may have a modest effect on immune response.


Diabetes
```{r}
p_diabetes = ggplot(dat1, aes(x = factor(diabetes), y = log_antibody)) +
  geom_boxplot(fill = "salmon") +
  labs(title = "Antibody Levels by Diabetes Status", x = "Diabetes (0 = No, 1 = Yes)", y = "log_antibody") +
  theme_minimal()

p_diabetes
```

* Very slight difference in medians.
* Distributions are largely overlapping.
* Diabetes might not have a strong effect on antibody levels in this sample.

Race
```{r}
p_race = ggplot(dat1, aes(x = factor(race), y = log_antibody)) +
  geom_boxplot(fill = "khaki") +
  labs(
    title = "Antibody Levels by Race",
    x = "Race (1 = White, 2 = Asian, 3 = Black, 4 = Hispanic)",
    y = "log_antibody"
  ) +
  theme_minimal()
p_race
```
* No substantial differences across race groups.
* Medians are nearly identical.
* Interpretation:
* Race may not be a strong predictor of antibody levels here.

Hypertension status
```{r}
p_htn = ggplot(dat1, aes(x = factor(hypertension), y = log_antibody)) +
  geom_boxplot(fill = "lightgreen") +
  labs(
    title = "Antibody Levels by Hypertension Status",
    x = "Hypertension (0 = No, 1 = Yes)",
    y = "log_antibody"
  ) +
  theme_minimal()
p_htn
```

* Slight difference: non-hypertensive individuals may have slightly higher antibody levels.
* Distributions are still very similar.

We will creat a panel including all our categrical predictors analysis:
```{r categorical-panel, fig.width=12, fig.height=8}
(p_gender | p_race | p_smoking) /
(p_diabetes | p_htn) 
```

4. Scatterplots with continuous predictors

BMI
```{r}
p_bmi = ggplot(dat1, aes(x = bmi, y = log_antibody)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "loess", se = FALSE, color = "darkgreen") +
  labs(title = "Antibody levels by BMI", x = "BMI", y = "log_antibody") +
  theme_minimal()

p_bmi
```
* Antibody levels are relatively flat until ~25–28 BMI, then slightly decline.
* Weak, nonlinear negative association at higher BMI.

Antibody vs. Age
```{r}
p_age = ggplot(dat1, aes(x = age, y = log_antibody)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "loess", se = FALSE, color = "blue") +
  labs(title = "Antibody Levels by Age", x = "Age (years)", y = "log_antibody") +
  theme_minimal()
p_age
```
* Slight decline in antibody levels as age increases.
* Smooth trend suggests a gradual, nonlinear effect.

Antibody vs. Systolic Blood Pressure (SBP)
```{r}
p_sbp = ggplot(dat1, aes(x = SBP, y = log_antibody)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "loess", se = FALSE, color = "darkred") +
  labs(title = "Antibody Levels by Systolic Blood Pressure", x = "SBP (mmHg)", y = "log_antibody") +
  theme_minimal()

p_sbp
```
* No clear trend — curve is almost flat across the SBP range.
* SBP appears to have minimal or no relationship with antibody levels.


Antibody vs. LDL Cholesterol
```{r}
p_ldl = ggplot(dat1, aes(x = LDL, y = log_antibody)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "loess", se = FALSE, color = "purple") +
  labs(title = "Antibody Levels by LDL Cholesterol", x = "LDL (mg/dL)", y = "log_antibody") +
  theme_minimal()

p_ldl
```

* Nearly flat curve — no apparent association.
* LDL likely not predictive of antibody response.

```{r continuous-panel, fig.width=16, fig.height=6}
(p_bmi | p_age) /
(p_sbp | p_ldl)
```


Next step, we choose the GAM model training for option 1

```{r}
mod_gam_full = gam(
  log_antibody ~ s(time) + s(age) + s(bmi) +
    LDL + SBP +
    gender + race + smoking + diabetes + hypertension,
  data = dat1
)

summary(mod_gam_full)      
plot(mod_gam_full, pages = 1)
```
All three smooth terms are highly significant (p < 2e-16)

Time, age, and BMI have nonlinear relationships with log_antibody

Their estimated degrees of freedom (edf):

* s(time) = 7.86 → fairly wiggly

* s(age) = 1.00 → almost linear

* s(bmi) = 5.13 → moderately nonlinear

Interpretation:Smooth terms for time, age, and BMI were statistically significant, indicating non-linear effects on antibody levels. The model automatically adapted flexibility for each predictor based on the data.

Model performance of the traning set
```{r}
# Predict log_antibody in the test set
pred_gam = predict(mod_gam_full, newdata = dat2)

# True observed values
true_values = dat2$log_antibody


# Create a tibble with true and predicted values
eval_df = tibble(
  truth = dat2$log_antibody,
  prediction = pred_gam
)

# Get multiple metrics at once
metrics(eval_df, truth = truth, estimate = prediction)
```
Using the GAM model on this external dataset, the model achieved a root mean squared error (RMSE) of 0.539 and a mean absolute error (MAE) of 0.433, indicating modest prediction error. 
The model explained approximately 17.1% of the variation in log-transformed antibody levels (R² = 0.171). While performance was not exceptionally high, this is consistent with the inherent complexity and noise in biological data. 
The relatively stable performance across the two datasets suggests that the model generalizes reasonably well to new individuals.
